---
title: "Birthrates in Swiss Districts"
author: "Sujal Manandhar"
output:
  pdf_document: default
  html_notebook: default
---

# Install and load the plgraphics package
```{r}
library(plgraphics)
library(ggplot2)
library(reshape2)
```

### Load data
```{r}
data = read.csv("swiss.csv")
data
```
```{r}
num_rows <- nrow(data)
num_column <- ncol(data)
print(num_rows)
print(num_column)
```

# Load the dataset
```{r}
data("d.birthrates")
```


### Data Cleaning and Preprocessing 

# Check for missing values
```{r}
missing_values <- colSums(is.na(d.birthrates))
print("Missing Values:")
print(missing_values)
```

### Exploratory Data Analysis (EDA)

# View the first few rows of the dataset
```{r}
head(d.birthrates)
```

# View the last few rows of the dataset
```{r}
tail(d.birthrates)
```

# Summary Statistics

Task: Start by examining summary statistics of the dataset to understand the distribution and range of each variable.

```{r}
# Assuming d.birthrates is your dataframe
plot(d.birthrates$fertility, type = "o", xlab = "Row Number", ylab = "Fertility", main = "Fertility Over Rows")
```

Correlation Analysis

Task: Calculate the correlation matrix to identify relationships between variables.

```{r}
# Calculate correlation matrix
correlation_matrix <- cor(d.birthrates[, c("fertility", "infantMort", "catholic", "single24", "single49")])

# Convert correlation matrix to long format for plotting
correlation_long <- melt(as.matrix(correlation_matrix))

# Plot heatmap with correlation values inside boxes
ggplot(correlation_long, aes(Var1, Var2, fill=value, label=round(value, 2))) +
  geom_tile(color="white") +
  scale_fill_gradient2(low="blue", high="red", mid="white", 
                       midpoint=0, limit=c(-1,1), space="Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1)) +
  geom_text(color="black", size=3) +
  coord_fixed()
```

# Visualization

Task: Create scatterplot matrices, boxplots, barplots, and histograms to visualize the relationships between variables and explore distributions.

# Scatterplot Matrix

```{r}
pairs(~fertility + infantMort + catholic + single24 + single49, data=d.birthrates, main="Scatterplot Matrix")
```

# Boxplot of fertility by language

```{r}
boxplot(fertility ~ language, data=d.birthrates, main="Fertility by Dominating Language", xlab="Language", ylab="Fertility")
```


# # Barplot of percentage of Catholic by altitude category

```{r}
barplot(tapply(d.birthrates$catholic, d.birthrates$altitude, mean), 
        main="Percentage of Catholic by Altitude Category",
        xlab="Altitude Category",
        ylab="Percentage Catholic")
```


# # Histogram of infant mortality

```{r}
hist(d.birthrates$infantMort, main="Histogram of Infant Mortality", xlab="Infant Mortality")
```

# Plot fertility against other variables
```{r}
par(mfrow=c(2,2))
plot(d.birthrates$fertility, d.birthrates$infantMort, xlab="Fertility", ylab="Infant Mortality", main="Fertility vs Infant Mortality")
plot(d.birthrates$fertility, d.birthrates$catholic, xlab="Fertility", ylab="Percentage Catholic", main="Fertility vs Percentage Catholic")
plot(d.birthrates$fertility, d.birthrates$single24, xlab="Fertility", ylab="Percentage Single (20-24)", main="Fertility vs Percentage Single (20-24)")
plot(d.birthrates$fertility, d.birthrates$single49, xlab="Fertility", ylab="Percentage Single (45-49)", main="Fertility vs Percentage Single (45-49)")
```


### Linear Regression

```{r}
library(caret)
library(lattice)
library(performance)
```

```{r}
# Split the data into training and testing sets
set.seed(123) # For reproducibility
train_index <- createDataPartition(d.birthrates$fertility, p = 0.8, list = FALSE)
train_data <- d.birthrates[train_index, ]
test_data <- d.birthrates[-train_index, ]

# Fit your model (e.g., linear regression)
model <- lm(fertility ~ infantMort + catholic + single24 + single49, data=d.birthrates)

# Check the fitted model
print(model)

# Check for influential points (Cook's distance plot)
cooksd <- cooks.distance(model)
plot(cooksd, pch = "*", cex = 1, main = "Influential Points by Cook's distance")
abline(h = 4*mean(cooksd, na.rm = TRUE), col = "red")  # Highlighting cutoff for influential points
```

```{r}
# 1. Residuals vs Fitted: Checking for homoscedasticity
plot(model, which = 1)

# 2. Normal Q-Q plot: Checking for normality of residuals
plot(model, which = 2)

# 3. Scale-Location plot: Another check for homoscedasticity
plot(model, which = 3)

# 4. Cook's distance plot: Identifying influential points
plot(model, which = 4)
```



### Evaluation

# Calculate MSE

```{r}
predicted <- predict(model, newdata=test_data)
mse <- mean((test_data$fertility - predicted)^2)
print(paste("Mean Squared Error (MSE):", mse))
```

# Visualize Results

```{r}
plot(test_data$fertility, predicted, main = "Actual vs Predicted Fertility", xlab = "Actual Fertility", ylab = "Predicted Fertility")
abline(0, 1, col = "red")

# Add MSE to the plot
legend("topleft", legend = paste("MSE:", round(mse, 2)), bty = "n")
```

# Summary of the model

```{r}
# Extract coefficients from the linear regression model
coefficients <- coef(model)[-1]  # Exclude intercept

# Plot coefficients
barplot(coefficients, main="Coefficients of Linear Regression Model", 
        ylab="Coefficient Value", names.arg=names(coefficients))
```



